Standartox: Standardising toxicity data

### 2019-12-19 article/parts/02_data_description.tex
Updates from Overleaf

\subsection{Accuracy Assessment}
Other initiatives harmonizing the large number of ecotoxicological test results were also developed recently. They partly aim for overlapping goals, yet have limitations or objectives that differentiate them from Standartox. Since 2006, the PPDB provides validated ecotoxicological information on pesticides \citep{lewis_international_2016}. In order to asses the validity of the Standartox aggregation approach, we compared Standartox results with effect values from the PPDB (n=3601) for the same chemicals (Figure \ref{fig:standartox_ppdb_diff}). This showed that 91.3\% of the Standartox values diverge by less than a factor of ten from the PPDB values. Considering only Standartox values for which at least five or ten tests were aggregated the value increases to 91.7\% and 93.4\%, respectively. Admittedly a few Standartox values diverge by a large margin.

@@

\subsection{Limitations}
Since the data in Standartox merely represents a sample of the actual toxicity of chemicals on organisms, it has to be stated that the returned Standartox values are subject to change. New studies will be published and constantly incorporated into EPA ECOTOX data base and therefore into Standartox, improving the toxicity estimates. However, we expect values of well tested chemical-organism combinations to be relatively stable. A future analysis on the smallest number of tests needed to return trustworthy results, could shed more light on this aspect. Furthermore, toxicity test results are certainly influenced by individual tests parameters, such as pH, temperature or conductivity amongst others \citep{rosenkrantz_influence_2013, li_temperature_2011}. Such information is not included in the aggregation performed by Standartox, since the EPA ECOTOX data base only provides sparse records on those. Most frequently provided test parameters (and their percentage) are medium temperature (77\%), pH:56\%, hardness:27\%, dissolved oxygen:18\%, Alkalinity:15\% and salinity:9\%. Others are provided for less than 5\% of the tests. A text-mining approach, iterating through the individual publications could potentially increase this number. However, we argue that using the geometric mean as an aggregate allows for an adequate estimation of the toxicity of a chemical towards an organism group. We chose the geometric mean in comparison to the arithmetic mean, since it is considerably less influenced by outliers and is suitable for skewed data. Also, the geometric mean is preferable over the median, since the median completely ignores the influence of large or small values, making it unreliable for small data sets <CITATION GEOMETRIC MEAN>.

>>>>

\subsection{Accuracy Assessment}
Other initiatives harmonizing the large number of ecotoxicological test results were also developed recently. They partly aim for overlapping goals, yet have limitations or objectives that differentiate them from Standartox. Since 2006, the PPDB provides validated ecotoxicological information on pesticides \citep{lewis_international_2016}. The aggregation of Standartox matches PPDB (n=3601) values that have been manually quality checked in the 91 \% within 1 order of magnitude. Considering only Standartox values for which at least ten tests were aggregated the value increases to 93.4 \%. Similarly, we compared Standartox to values from the ChemProp \citep{ufzdepartmentofecologicalchemistry_chemprop_2019} software, which calculates LC\textsubscript{50} values by the Read Across method \citep{schuurmann_quantitative_2011} and found that (n=179) 95 \% lie within a factor 10 difference. However, the difference is not necessarily an indication of lower quality of these data in Standartox but may also be attributed to the inclusion of consideration of a wider range of test data and conditions as well as variance in the Read Across models, respectively.

In order to asses the validity of the Standartox aggregation approach, we compared Standartox results with effect values from the PPDB (n=3601) for the same chemicals (Figure \ref{fig:standartox_ppdb_diff}).

  Admittedly a few Standartox values diverge by a large margin.

@@

\citet{vandenberg_modeling_2019} used ECOTOX for trait-based sensitivity models. \include{article/values/publish_rate.tex}

\subsection{Outlook}
Standartox is not a one point in time effort, but an automated tool that steadily incorporates newly published ecotoxicity data. The ECOTOX database added on average \verbatiminput{article/values/test.txt} new test results in the five last years.

Standartox is not a singular proposition of a tool that aggregates large ecotoxicity data, but constitutes an automated tool that

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% CONTINUE HERE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%55

Since the data in Standartox merely represents a sample of the actual toxicity of chemicals on organisms, it has to be stated that the returned Standartox values are subject to change. New studies will be published and constantly incorporated into EPA ECOTOX data base and therefore into Standartox, improving the toxicity estimates. However, we expect values of well tested chemical-organism combinations to be relatively stable. A future analysis on the smallest number of tests needed to return trustworthy results, could shed more light on this aspect. Furthermore, toxicity test results are certainly influenced by individual tests parameters, such as pH, temperature or conductivity amongst others \citep{rosenkrantz_influence_2013, li_temperature_2011}. Such information is not included in the aggregation performed by Standartox, since the EPA ECOTOX data base only provides sparse records on those. Most frequently provided test parameters (and their percentage) are medium temperature (77\%), pH:56\%, hardness:27\%, dissolved oxygen:18\%, Alkalinity:15\% and salinity:9\%. Others are provided for less than 5\% of the tests. A text-mining approach, iterating through the individual publications could potentially increase this number. However, we argue that using the geometric mean as an aggregate allows for an adequate estimation of the toxicity of a chemical towards an organism group. We chose the geometric mean in comparison to the arithmetic mean, since it is considerably less influenced by outliers and is suitable for skewed data. Also, the geometric mean is preferable over the median, since the median completely ignores the influence of large or small values, making it unreliable for small data sets <CITATION GEOMETRIC MEAN>.
